---
title: "Homework week 05"
author: "Daniel"
date: "2025-03-22"
output: html_document
---

# Compute the regression coefficients for fish = b0+b1*copepod
# use randomization method to generate null distribution of b1
# test whether b1 is significantly different from null (with randomization 5000 times). 
# Report your p-value.

```{r}
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data")
library(readxl)
f_c_den <- read_excel("mock_dataTS/enviANDdensity.xls")
fish_den <- f_c_den$`FishDensity (ind./1000m3)`
copepod_den <- f_c_den$`CopepodDensity ind./m3)`
```

### calculate b1 of original data

```{r}
x <- cbind(1, copepod_den)
y <- fish_den
b <- solve(t(x) %*% x) %*% t(x) %*% y
b1 <- b[2]
print(b1)
```

### randomization

```{r}
n_p <- 5000
l <- length(fish_den)
b_null <- matrix(0, nrow = n_p, ncol = 2)
b1_null <- numeric(n_p)
fish_den_shuffle <- numeric(n_p)

set.seed(123)
for (i in 1:n_p){
  fish_den_shuffle <- fish_den[order(runif(l))] #make an random order of 1 to 34
  b_null[i,] <- solve(t(x) %*% x) %*% t(x) %*% fish_den_shuffle #solve b by the shuffle one
  b1_null[i] <- b_null[i, 2] #extract every b1 into b1_null
}
```

### calculate p-value

```{r}
p1 <- 2 * min(mean(b1_null >= b1), mean(b1_null <= b1))

if(p1 < (1/5000)){
  print(paste("p-value < 1/5000"))
} else{
  print(paste("p-value:", p1))
}

if(p1 < 0.05){
  print("b1 and 0 are significantly different")
} else{
  print("b1 and 0 are not significantly different")
}
```

# Randomization test whether significant difference exists between the density of Oncaea Venusta and Canthocalanus pauper. 
# (Assume all data are independent and use all 34 stations.) 
# Report your p-value.

```{r}
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data")
cop_compo <- read.table("mock_dataTS/copepod_composition.txt", header = T)
cop_den <- read.table("mock_dataTS/cop_density.txt")
cop_sp <- read.table("mock_dataTS/copepodSPlist.txt", fill = T, sep = "\n")
```

### calculate th density of these two species

```{r}
# same as before
row.names(cop_compo) <- cop_sp$V1
Onve <- cop_compo["Oncaea venusta",] / 100
Capa <- cop_compo["Canthocalanus pauper",] / 100

Onve_den <- matrix(0, nrow = 1, ncol = length(cop_den$V1))
row.names(Onve_den) <- "Oncaea venusta"
for(i in 1:length(cop_den$V1)){
  Onve_den[, i] <- Onve[, i] * cop_den[i,]
}

Capa_den <- matrix(0, nrow = 1, ncol = length(cop_den$V1))
row.names(Capa_den) <- "Canthocalanus pauper"
for(i in 1:length(cop_den$V1)){
  Capa_den[, i] <- Capa[, i] * cop_den[i,]
}
```

### calculate the difference of original data

```{r}
original_diff <- mean(Onve_den) - mean(Capa_den)
print(original_diff)
```

### randomization

```{r}
L <- length(Onve_den)
diff_null <-numeric(n_p)

set.seed(123)
for(i in 1:n_p){
  #combine two species and shuffle, then seperate into two groups
  combine <- c(Onve_den, Capa_den) 
  com_shuffle <- combine[order(runif(length(combine)))] #make an random order from 1 to 68
  Onve_shuffle <- com_shuffle[1:length(Onve_den)] #first half for Onve
  Capa_shuffle <- com_shuffle[(length(Onve_den) + 1):length(com_shuffle)] #second half for Capa
  
  diff_null[i] <- mean(Onve_shuffle) - mean(Capa_shuffle) #calculate the null difference
}
```

### calculate the p-value

```{r}
p2 <- 2 * min(mean(diff_null <= original_diff), mean(diff_null >= original_diff))

if(p2 < (1/5000)){
  print(paste("p-value < 1/5000"))
} else{
  print(paste("p-value:", p2))
}

if(p2 < 0.05){
  print("these two species are significantly different")
} else{
  print("these two species are not significantly different")
}
```
