---
title: "Homework week 09"
author: "Daniel"
date: "2025-04-18"
output: html_document
---

* Use the dominant copepod species data (from HW 1). 
* Perform cluster analysis of stations based on percent composition data of the dominant species and tell your story about these copepod data. 
* Compare the results based on different distance measures. 
* Compare the results based on different cluster algorithms. 
* Determine final number of clusters and describe the differences among them.

dominant species:
```{r}
#same as Hw01
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data/mock_dataTS")
cop_compo <- read.table("Copepod_composition.txt", header = T)
cop_den <- read.table("Cop_density.txt")
cop_sp <- read.table("copepodSPlist.txt", fill = T, sep = "\n")

sp_den <- matrix(0, nrow = length(cop_sp$V1), ncol = length(cop_den$V1))
for(i in 1:length(cop_sp$V1)){
  for(j in 1:length(cop_den$V1)){
    sp_den[i, j] <- cop_compo[i, j]/100 * cop_den[j,]
  }
}
sp_den.df <- as.data.frame(sp_den)
colnames(sp_den.df) <- colnames(cop_compo)
rownames(sp_den.df) <- cop_sp$V1
head(sp_den.df)

dominant_sp.df <- data.frame(ifelse(cop_compo >= 5, 1, 0))
dominant_species <- rownames(dominant_sp.df)[rowSums(dominant_sp.df) >= 1]
as.numeric(dominant_species)

dom_sp_compo <- t(cop_compo[dominant_species, ]) #exchange x-axis and y-axis to compare with different cruise station
colnames(dom_sp_compo) <- cop_sp[dominant_species, ] #change the column names into dominant species names
rownames(dom_sp_compo) <- colnames(cop_compo)
```

```{r}
library(vegan)
library(cluster)
```

Comparing with four different distance measures

1. Euclidean
```{r}
dist_eu <- dist(dom_sp_compo, "euclidean")

hc_eu_average <- hclust(dist_eu, "average") #use the hclust function to make hierarchical cluster by average method
hc_eu_ward <- hclust(dist_eu, "ward.D2") #by modified ward method

par(mfrow=c(1,2))

plot(hc_eu_average, main = "Euclidean + Average")
cutree(hc_eu_average, k = 8) #manually determine into 8 clusters
rect.hclust(hc_eu_average, k = 8) #plot the number of clusters on the hierarchical cluster

plot(hc_eu_ward, main = "Euclidean + Ward")
cutree(hc_eu_ward, k = 3)
rect.hclust(hc_eu_ward, k = 3)
```

2. Manhattan
```{r}
# same as above
dist_man <- dist(dom_sp_compo, "manhattan")

hc_man_average <- hclust(dist_man, "average")
hc_man_ward <- hclust(dist_man, "ward.D2")

par(mfrow=c(1,2))

plot(hc_man_average, main = "Manhattan + Average")
cutree(hc_man_average, k = 3)
rect.hclust(hc_man_average, k = 3)

plot(hc_man_ward, main = "Manhattan + Ward")
cutree(hc_man_ward, k = 3)
rect.hclust(hc_man_ward, k = 3)
```

3. Bray-Crutis
```{r}
dist_bc <- vegdist(dom_sp_compo, "bray")

hc_bc_average <- hclust(dist_bc, "average")
hc_bc_ward <- hclust(dist_bc, "ward.D2")

par(mfrow=c(1,2))

plot(hc_bc_average, main = "Bray-Curtis + Average")
cutree(hc_bc_average, k = 3)
rect.hclust(hc_bc_average, k = 3)

plot(hc_bc_ward, main = "Bray-Curtis + Ward")
cutree(hc_bc_ward, k = 3)
rect.hclust(hc_bc_ward, k = 3)
```

4. Jaccard
```{r}
dist_ja <- vegdist(dom_sp_compo, "jaccard", binary = T)

hc_ja_average <- hclust(dist_ja, "average")
hc_ja_complete <- hclust(dist_ja, "complete")

par(mfrow=c(1,2))

plot(hc_ja_average, main = "Jaccard + Average")
cutree(hc_ja_average, k = 4)
rect.hclust(hc_ja_average, k = 4)

plot(hc_ja_complete, main = "Jaccard + Complete")
cutree(hc_ja_complete, k = 4)
rect.hclust(hc_ja_complete, k = 4)
```

```{r}
distance <- c("Euclidean", "Manhattan", "Bray-Curtis", "Jaccard")
k_average <- c(8, 3, 3, 4)
k_ward <- c(3, 3, 3, NA)
k_complete <- c(NA, NA, NA, 4)
k_compare <- data.frame(distance, k_average, k_ward, k_complete)
k_compare
```

Cophenetic Correlation
```{r}
coph_eu_average <- round(cor(cophenetic(hc_eu_average), dist_eu), 3)
coph_eu_ward <- round(cor(cophenetic(hc_eu_ward), dist_eu), 3)
coph_man_average <- round(cor(cophenetic(hc_man_average), dist_man), 3)
coph_man_ward <- round(cor(cophenetic(hc_man_ward), dist_man), 3)
coph_bc_average <- round(cor(cophenetic(hc_bc_average), dist_bc), 3)
coph_bc_ward <- round(cor(cophenetic(hc_bc_ward), dist_bc), 3)
coph_ja_average <- round(cor(cophenetic(hc_ja_average), dist_ja), 3)
coph_ja_complete <- round(cor(cophenetic(hc_ja_complete), dist_ja), 3)
```

```{r}
distance <- c("Euclidean", "Manhattan", "Bray-Curtis", "Jaccard")
coph_average <- c(coph_eu_average, coph_man_average, coph_bc_average, coph_ja_average)
coph_ward <- c(coph_eu_ward, coph_man_ward, coph_bc_ward, NA)
coph_complete <- c(NA, NA, NA, coph_ja_complete)
cop_compare <- data.frame(distance, coph_average, coph_ward, coph_complete)
cop_compare
```

non-Hierarchical

1. Euclidean
```{r}
sil_eu <- sapply(2:8, function(k){ #testing different k from 2 to 8
  pam_eu <- pam(dist_eu, k)
  pam_eu$silinfo$avg.width #showing average silhouette width of 2 clusters to 8 clusters
})

par(mfrow=c(1,1))
plot(2:8, sil_eu, type = "b", #plot the average silhouette width to find the best number of clusters
     xlab = "k",
     ylab = "Average Silhouette Width",
     main = "Euclidean")

pam_eu_best <- pam(dist_eu, k = 3)
pam_eu_best$clustering #showing all the stations belong to which cluster

plot(silhouette(pam_eu_best), border = NA, main = "k = 3") #separate those stations into three clusters
```

2. Manhattan
```{r}
sil_man <- sapply(2:8, function(k){
  pam_man <- pam(dist_man, k)
  pam_man$silinfo$avg.width
})

par(mfrow=c(1,1))
plot(2:8, sil_man, type = "b",
     xlab = "k",
     ylab = "Average Silhouette Width",
     main = "Manhattan")

pam_man_best <- pam(dist_man, k = 3)
pam_man_best$clustering

plot(silhouette(pam_man_best), border = NA, main = "k = 3")
```

3. Bray-Crutis
```{r}
sil_bc <- sapply(2:8, function(k){
  pam_bc <- pam(dist_bc, k)
  pam_bc$silinfo$avg.width
})

par(mfrow=c(1,1))
plot(2:8, sil_bc, type = "b",
     xlab = "k",
     ylab = "Average Silhouette Width",
     main = "Bray-Curtis")

pam_bc_best <- pam(dist_bc, k = 3)
pam_bc_best$clustering

plot(silhouette(pam_bc_best), border = NA, main = "k = 3")
```

4. Jaccard
```{r}
sil_ja <- sapply(2:8, function(k){
  pam_ja <- pam(dist_ja, k)
  pam_ja$silinfo$avg.width
})

par(mfrow=c(1,1))
plot(2:8, sil_ja, type = "b",
     xlab = "k",
     ylab = "Average Silhouette Width",
     main = "Jaccard")

pam_ja_best <- pam(dist_ja, k = 2)
pam_ja_best$clustering

plot(silhouette(pam_ja_best), border = NA, main = "k = 2")
```

```{r}
distance <- c("Euclidean", "Manhattan", "Bray-Curtis", "Jaccard")
best_k <- c(3, 3, 3, 2)
average_silhouette <- c(0.27, 0.28, 0.29, 0.44)
compare <- data.frame(distance, best_k, average_silhouette)
compare
```

1. From k_compare, Euclidean + average linkage resulted in a chaining pattern, leading to the choice of 8 clusters, which contrasts with the more consistent 3-cluster solutions from Ward linkage across Euclidean, Manhattan, and Bray-Curtis distances.
2. Form cop_compare, the dendrogram constructed using average linkage based on Jaccard distance most faithfully preserved the original dissimilarity structure.
3. From "compare", Jaccard distance (binary) performed best with an average silhouette width of 0.44 when using 2 clusters.
  + In contrast, abundance-based distances such as Bray-Curtis, Manhattan, and Euclidean all selected 3 clusters but had significantly lower silhouette values (0.27–0.29), suggesting weaker group separation.
  + This result indicates that presence/absence differences among stations may be more important than fine-scale abundance variation in structuring the copepod community across seasons.

