---
title: "Homework week 09"
author: "Daniel"
date: "2025-04-18"
output: html_document
---

## Use the dominant copepod species data (from HW 1). 
## Perform cluster analysis of stations based on percent composition data of the dominant species and tell your story about these copepod data. 
## Compare the results based on different distance measures. 
## Compare the results based on different cluster algorithms. 
## Determine final number of clusters and describe the differences among them.

### dominant species
```{r}
#same as Hw01
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data/mock_dataTS")
cop_compo <- read.table("Copepod_composition.txt", header = T)
cop_den <- read.table("Cop_density.txt")
cop_sp <- read.table("copepodSPlist.txt", fill = T, sep = "\n")

sp_den <- matrix(0, nrow = length(cop_sp$V1), ncol = length(cop_den$V1))
for(i in 1:length(cop_sp$V1)){
  for(j in 1:length(cop_den$V1)){
    sp_den[i, j] <- cop_compo[i, j]/100 * cop_den[j,]
  }
}
sp_den.df <- as.data.frame(sp_den)
colnames(sp_den.df) <- colnames(cop_compo)
rownames(sp_den.df) <- cop_sp$V1
head(sp_den.df)

dominant_sp.df <- data.frame(ifelse(cop_compo >= 5, 1, 0))
dominant_species <- rownames(dominant_sp.df)[rowSums(dominant_sp.df) >= 1]
as.numeric(dominant_species)

dom_sp_compo <- t(cop_compo[dominant_species, ]) #exchange x-axis and y-axis to compare with different cruise station
colnames(dom_sp_compo) <- cop_sp[dominant_species, ] #change the column names into dominant species names
```

```{r}
library(vegan)
library(cluster)
```

### Comparing with four different distance measures
#### Euclidean
```{r}
dist_eu <- dist(dom_sp_compo, "euclidean") #calculate the euclidean distance

sil_eu <- sapply(2:8, function(k){ #testing different k from 2 to 8
  pam_eu <- pam(dist_eu, k) #use function pam to cluster with different k
  pam_eu$silinfo$avg.width #giving the average silhouette width
})

par(mfrow=c(1,1))
#plot to find the highest average silhouette width
plot(2:8, sil_eu, type = "b",
     xlab = "k",
     ylab = "Average Silhouette Width",
     main = "Euclidean")
```

```{r}
pam_eu_best <- pam(dist_eu, k = 3) #k = 3 has the highest
pam_eu_best$clustering #showing all stations belongs to which cluster

plot(silhouette(pam_eu_best), border = NA, main = "k = 3") #plot by the best cluster number
pam_eu_best$silinfo$avg.width #the average of all silhouette width

hc_eu_average <- hclust(dist_eu, "average") #use the hclust function to make hierarchical cluster by average method
hc_eu_ward <- hclust(dist_eu, "ward.D2") #by modified ward method

par(mfrow=c(1,2))
plot(hc_eu_average, main = "Euclidean + Average") #plot the hierarchical cluster
rect.hclust(hc_eu_average, k = 3, border = "red") #plot the best number of clusters on the hierarchical cluster

plot(hc_eu_ward, main = "Euclidean + Ward")
rect.hclust(hc_eu_ward, k = 3, border = "red")
```

#### Manhattan
```{r}
#same as above
dist_man <- dist(dom_sp_compo, "manhattan")

sil_man <- sapply(2:8, function(k){
  pam_man <- pam(dist_man, k)
  pam_man$silinfo$avg.width
})

par(mfrow=c(1,1))
plot(2:8, sil_man, type = "b",
     xlab = "k",
     ylab = "Average Silhouette Width",
     main = "Manhattan")

pam_man_best <- pam(dist_man, k = 3)
pam_man_best$clustering

plot(silhouette(pam_man_best), border = NA, main = "k = 3")
pam_man_best$silinfo$avg.width

hc_man_average <- hclust(dist_man, "average")
hc_man_ward <- hclust(dist_man, "ward.D2")

par(mfrow=c(1,2))
plot(hc_man_average, main = "Manhattan + Average")
rect.hclust(hc_man_average, k = 3, border = "red")

plot(hc_man_ward, main = "Manhattan + Ward")
rect.hclust(hc_man_ward, k = 3, border = "red")
```

#### Bray-Curtis
```{r}
dist_bc <- vegdist(dom_sp_compo, "bray")

sil_bc <- sapply(2:8, function(k){
  pam_bc <- pam(dist_bc, k)
  pam_bc$silinfo$avg.width
})

par(mfrow=c(1,1))
plot(2:8, sil_bc, type = "b",
     xlab = "k",
     ylab = "Average Silhouette Width",
     main = "Bray-Curtis")

pam_bc_best <- pam(dist_bc, k = 3)
pam_bc_best$clustering

plot(silhouette(pam_bc_best), border = NA, main = "k = 3")
pam_bc_best$silinfo$avg.width

hc_bc_average <- hclust(dist_bc, "average")
hc_bc_ward <- hclust(dist_bc, "ward.D2")

par(mfrow=c(1,2))
plot(hc_bc_average, main = "Bray-Curtis + Average")
rect.hclust(hc_bc_average, k = 3, border = "red")

plot(hc_bc_ward, main = "Bray-Curtis + Ward")
rect.hclust(hc_bc_ward, k = 3, border = "red")
```

#### Jaccard
```{r}
dist_ja <- vegdist(dom_sp_compo, "jaccard", binary = T)

sil_ja <- sapply(2:8, function(k){
  pam_ja <- pam(dist_ja, k)
  pam_ja$silinfo$avg.width
})

par(mfrow=c(1,1))
plot(2:8, sil_ja, type = "b",
     xlab = "k",
     ylab = "Average Silhouette Width",
     main = "Jaccard")

pam_ja_best <- pam(dist_ja, k = 2)
pam_ja_best$clustering

plot(silhouette(pam_ja_best), border = NA, main = "k = 2")
pam_ja_best$silinfo$avg.width

hc_ja_average <- hclust(dist_ja, "average")
hc_ja_complete <- hclust(dist_ja, "complete")

par(mfrow=c(1,2))
plot(hc_ja_average, main = "Jaccard + Average")
rect.hclust(hc_ja_average, k = 2, border = "red")

plot(hc_ja_complete, main = "Jaccard + Ward")
rect.hclust(hc_ja_complete, k = 2, border = "red")
```

### Compare different method and result
```{r}
par(mfrow=c(2,4))
plot(hc_eu_average, main = "Euclidean + Average")
rect.hclust(hc_eu_average, k = 3, border = "red")

plot(hc_eu_ward, main = "Euclidean + Ward")
rect.hclust(hc_eu_ward, k = 3, border = "red")

plot(hc_man_average, main = "Manhattan + Average")
rect.hclust(hc_man_average, k = 3, border = "red")

plot(hc_man_ward, main = "Manhattan + Ward")
rect.hclust(hc_man_ward, k = 3, border = "red")

plot(hc_bc_average, main = "Bray-Curtis + Average")
rect.hclust(hc_bc_average, k = 3, border = "red")

plot(hc_bc_ward, main = "Bray-Curtis + Ward")
rect.hclust(hc_bc_ward, k = 3, border = "red")

plot(hc_ja_average, main = "Jaccard + Average")
rect.hclust(hc_ja_average, k = 2, border = "red")

plot(hc_ja_complete, main = "Jaccard + Ward")
rect.hclust(hc_ja_complete, k = 2, border = "red")
```

```{r}
cluster_compare <- data.frame(
  Euclidean = pam_eu_best$clustering,
  Manhattan = pam_man_best$clustering,
  BrayCurtis = pam_bc_best$clustering,
  Jaccard = pam_ja_best$clustering
)
cluster_compare
```

