---
title: "Homework week 10"
author: "Daniel"
date: "2025-04-28"
output: html_document
---

```{r}
#dominant species
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data/mock_dataTS")
cop_compo <- read.table("Copepod_composition.txt", header = T)
cop_den <- read.table("Cop_density.txt")
cop_sp <- read.table("copepodSPlist.txt", fill = T, sep = "\n")

sp_den <- matrix(0, nrow = length(cop_sp$V1), ncol = length(cop_den$V1))
for(i in 1:length(cop_sp$V1)){
  for(j in 1:length(cop_den$V1)){
    sp_den[i, j] <- cop_compo[i, j]/100 * cop_den[j,]
  }
}
sp_den.df <- as.data.frame(sp_den)
colnames(sp_den.df) <- colnames(cop_compo)
rownames(sp_den.df) <- cop_sp$V1
head(sp_den.df)

dominant_sp.df <- data.frame(ifelse(cop_compo >= 5, 1, 0))
dominant_species <- rownames(dominant_sp.df)[rowSums(dominant_sp.df) >= 1]

dom_sp_compo <- t(cop_compo[dominant_species, ])
colnames(dom_sp_compo) <- cop_sp[dominant_species, ]
rownames(dom_sp_compo) <- colnames(cop_compo)
dom_sp_compo.df <- as.data.frame(dom_sp_compo) #save the result of dominant species as data frame
```

Apply PCA or MDS using matrix algebra to extract major gradients of the dominant species. 

Make a bi-plot to show the relationships among species and sites. 

Then, check your results with those from build in functions of PCA, MDS, and NMDS.

* PCA manually
```{r}
dom_sp_center <- scale(dom_sp_compo.df, center = T, scale = F) #centering by subtracting the column means

cov_matrix <- cov(dom_sp_center) #build up a matrix contain the variance and the covariance
eigen_decompo <- eigen(cov_matrix) #compute the eigenvalue and eigenvector
eigen_vector <- eigen_decompo$vectors
eigen_value <- eigen_decompo$values

pca_score <- dom_sp_center %*% eigen_vector #project the centralized data frame on the new eigenvectors 

plot(pca_score[, 1], pca_score[, 2], #the first column as x-axis(PC1), and second column as y-axis(PC2), draw the scatter plot
     xlab = "PC1", 
     ylab = "PC2", 
     main = "Biplot")

text(pca_score[,1], pca_score[,2], #text the station name on the scatter plot
     labels = rownames(dom_sp_compo.df),
     pos = 4, cex = 0.8) #pos = 4, means the text position to the right of the spot, and cex change the size of the text

for (i in 1:ncol(dom_sp_compo.df)) { #draw all the arrows
  arrows(0, 0,  #started from (0, 0)
         eigen_vector[i, 1] * sqrt(eigen_value[1]) * 2, #loading * the square root of eigenvalue, * 2 scale for visualization
         eigen_vector[i, 2] * sqrt(eigen_value[2]) * 2,
         col = "blue", length = 0.1)
  
  text(eigen_vector[i, 1] * sqrt(eigen_value[1]) * 1.8 * 1.2, #define the position of dominant species names
       eigen_vector[i, 2] * sqrt(eigen_value[2]) * 1.8 * 1.2, 
       labels = colnames(dom_sp_compo.df)[i], 
       col = "blue", cex = 0.8)
}
```

* PCA (automatically)
```{r}
pca_dom_copepod <- prcomp(dom_sp_center)
biplot(pca_dom_copepod, scale = 0) #scale = 0, no adjustment of the arrow length
```

1. By comparing the manually generated and built-in biplots, it was observed that the relative distribution of sample points was consistent, and the directions of principal components were aligned (except for possible sign differences due to mathematical indeterminacy).

2. In the biplot, due to visual scaling of the species arrows, variable importance should be interpreted based on loading values rather than arrow lengths.