---
title: "Hw 02"
author: "Daniel"
date: "2025-03-04"
output: html_document
---

# 1a. Generate 10000 random numbers from Gaussian distribution with mean=20 and variance =10, and plot the distribution.
```{r}
set.seed(123)
g_dis <- rnorm(10000, 20, 10) #use the rnorm function to generate a Gaussian distribution

hist(g_dis) #plot the histogram
```

# 1b. Generate 10000 random numbers from Binomial distribution with p=0.5 and n=40, and plot the distribution.
```{r}
set.seed(123)
n <- 40 #setting n=40
p <- 0.5 #setting p=0.5
size <- 10000 #to generate 10000 numbers

b_dis <- rep(0, size) #make an empty numeric to store the result

for(i in 1:size){
  b_dis[i] <- sum(runif(n) < p) #generate 40 random numbers from uniform distribution and if the resulting number is smaller than 0.5, defined as success, then sum up the number of success to become the result of binomial distribution
}

hist(b_dis) #plot the histogram
```

# Compare the distribution of 1a and 1b, what do you find?
```{r}
par(mfrow = c(1, 2)) #change the default to compare two plots at the same time
hist(g_dis)
hist(b_dis)
```

### 1. Gaussian distribution has a wider result range
### 2. the result frequency is higher in binomial distribution relative to Gaussian distribution

# 2. Make a program that can select our candidates for presentation next week. This program should select randomly but avoid selecting the numbers that had been selected before.
```{r}
candi <- function (m, w){ #name the function "candi" and the arguments are the number of the class member "m", and total weeks of this semester"w"
  selected <- numeric(0) #build a empty numeric to record the number has been selected
  candidate <- matrix(0, nrow = w, ncol = 2) #generate an empty matrix to store the result
  
  for(week in 1:w){
    repeat{ #repeat the loop till all requirements are satisfied
      num1 <- round(runif(1, 1, m)) #use "round" to change the result of "runif" into integer
      num2 <- round(runif(1, 1, m))
      
      if(num1 != num2 && !(num1 %in% selected) && !(num2 %in% selected)){ #if num1 is unequal to num2 and both of them haven't been selected yet
        candidate[week, ] <- c(num1, num2) #then combine those two number into the candidate matrix
        selected <- c(selected, num1, num2) #also put those two number into the "selected" numeric
        break #break the loop and continue for next week
      }
    }
  }
  return(candidate) #return the result
}
```
```{r}
set.seed(123)
candi(34, 16)
```

