---
title: "Hw 02"
author: "Daniel"
date: "2025-03-04"
output: html_document
---

# 1a. Generate 10000 random numbers from Gaussian distribution with mean=20 and variance =10, and plot the distribution.
```{r}
set.seed(123)
g_dis1 <- rnorm(10000) #use the rnorm function to generate a Gaussian distribution
g_dis <- g_dis1  * 10 + 20 #change the standard deviation and mean
mean(g_dis)
sd(g_dis)
hist(g_dis) #plot the histogram
```

# 1b. Generate 10000 random numbers from Binomial distribution with p=0.5 and n=40, and plot the distribution.
```{r}
set.seed(123)
n <- 40 #setting n=40
p <- 0.5 #setting p=0.5
size <- 10000 #to generate 10000 numbers

b_dis <- rep(0, size) #make an empty numeric to store the result

for(i in 1:size){
  b_dis[i] <- sum(runif(n) < p) #generate 40 random numbers from uniform distribution and if the resulting number is smaller than 0.5, defined as success, then sum up the number of success to become the result of binomial distribution
}

hist(b_dis) #plot the histogram
```

# Compare the distribution of 1a and 1b, what do you find?
```{r}
par(mfrow = c(1, 2)) #change the default to compare two plots at the same time
hist(g_dis)
hist(b_dis)
```

### 1. Gaussian distribution has a wider result range
### 2. the result frequency is higher in binomial distribution relative to Gaussian distribution

# 2. Make a program that can select our candidates for presentation next week. This program should select randomly but avoid selecting the numbers that had been selected before.
```{r}
candi <- function(m){ #name the function "candi" with a argument "m" represent the number of students
  if (!exists("used", envir = .GlobalEnv)) { #if the numeric "used" is not exist, then initialize. ".GlobalEnv" make sure "used" will stay in the environment all time, and prevent it disappear after execute this function
    assign("used", numeric(0), envir = .GlobalEnv) #else assign "used" as a empty numeric to stored the number which has been chosen
  }  
  available <- setdiff(1:m, used) #use "setdiff" to select the number in 1:m but not in "used" and define them as "available"
  
  x <- runif(length(available)) #use the length of "available" to generate a uniform distribution
  candidate <- available[order(x)][1:2] #order them so the results are integer and take the first two as candidate
  
  assign("used", c(used, candidate), envir = .GlobalEnv) #assign "used" again but add those candidate into "used" 
  
  return(candidate) #return the result
}
```
```{r}
set.seed(123)
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
print(candi(30))
```
