---
title: "Homework week 13"
author: "Daniel"
date: "2025-05-22"
output: html_document
---

```{r}
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data")
corvina <- read.csv("HW_data2.csv")
age <- corvina$Age
length <- corvina$Length
```

### Please estimate the model parameters (growth coefficients and sigma) of four different growth functions (von Bertalanffy ; Gompertz ; Logistic; Schnute Richards) using the MLE method for Gulf Corvina ( Cynoscion othonopterus ) ), assuming a log normal error distribution.

#### VBGF
```{r}
#same as last week
nll_vbgf <- function(par, age, length){
  Lmax <- par[1]
  K <- par[2]
  t0 <- par[3]
  sigma <- par[4]
  
  L <- Lmax * (1 - exp(-K * (age - t0)))
  L[L <= 0] <- 1e-6
  
  n <- length(length)
  nll_value <- n * log(sigma) + sum((log(length) - log(L)) ^ 2) / (2 * sigma ^ 2)
  
  return(nll_value)
}

init_vbgf  <- c(680, 0.2, 0, 0.1)
lower_vbgf <- c(10, 0.001, -10, 0.0001)
upper_vbgf <- c(1000, 1, 10, 10)

fit_vbgf <- optim(par = init_vbgf, fn = nll_vbgf, age = age, length = length, 
                  method = "L-BFGS-B", lower = lower_vbgf, upper = upper_vbgf)

cat("NLL of VGBF = ", fit_vbgf$value, 
    "\nL∞ = ", fit_vbgf$par[1], "\nK = ", fit_vbgf$par[2], 
    "\nt0 = ", fit_vbgf$par[3], "\nσ ^ 2 = ", fit_vbgf$par[4] ^ 2)
```

#### Gompertz
```{r}
nll_gom <- function(par, age, length){
  Lmax <- par[1]
  k2 <- par[2]
  t0 <- par[3]
  sigma <- par[4]
  
  L <- Lmax * exp(-exp(-k2 * (age - t0)))
  L[L <= 0] <- 1e-6
  
  n <- length(length)
  nll_value <- n * log(sigma) + sum((log(length) - log(L)) ^ 2) / (2 * sigma ^ 2)
  
  return(nll_value)
}

init_gom  <- c(850, 0.5, 0, 0.1)
lower_gom <- c(120, 0.01, -5, 0.001)
upper_gom <- c(880, 5, 5, 1)

fit_gom <- optim(par = init_gom, fn = nll_gom, age = age, length = length, 
                 method = "L-BFGS-B", lower = lower_gom, upper = upper_gom)

cat("NLL of Gompertz = ", fit_gom$value, 
    "\nL∞ = ", fit_gom$par[1], "\nk2 = ", fit_gom$par[2], 
    "\nt0 = ", fit_gom$par[3], "\nσ ^ 2 = ", fit_gom$par[4] ^ 2)
```

#### Logistic Growth Function
```{r}
nll_log <- function(par, age, length){
  Lmax <- par[1]
  k3 <- par[2]
  t3 <- par[3]
  sigma <- par[4]
  
  L <- Lmax / (1 + exp(-k3 * (age - t3)))
  L[L <= 0] <- 1e-6
  
  n <- length(length)
  nll_value <- n * log(sigma) + sum((log(length) - log(L)) ^ 2) / (2 * sigma ^ 2)
  
  return(nll_value)
}

init_log  <- c(850, 0.5, 1, 0.1)
lower_log <- c(120, 0.01, -5, 0.001)
upper_log <- c(880, 5, 5, 1)

fit_log <- optim(par = init_log, fn = nll_log, age = age, length = length, 
                 method = "L-BFGS-B", lower = lower_log, upper = upper_log)

cat("NLL of Logistic Growth Function = ", fit_log$value, 
    "\nL∞ = ", fit_log$par[1], "\nk3 = ", fit_log$par[2], 
    "\nt3 = ", fit_log$par[3], "\nσ ^ 2 = ", fit_log$par[4] ^ 2)
```

#### Schnute-Richards
```{r}
nll_sch <- function(par, age, length){
  Lmax <- par[1]
  a <- par[2]
  k4 <- par[3]
  c <- par[4]
  b <- par[5]
  sigma <- par[6]
  
  L <- Lmax * (1 + a * exp(-k4 * age ^ c)) ^ (1 / b)
  L[L <= 0] <- 1e-6
  
  n <- length(length)
  nll_value <- n * log(sigma) + sum((log(length) - log(L)) ^ 2) / (2 * sigma ^ 2)
  
  return(nll_value)
}

init_sch <- c(705, 0.25, 0.1, 2.5, -0.15, 0.1)
lower_sch <- c(680, 0.01, 0.01, 0.1, -2, 0.001)
upper_sch <- c(1300, 10, 1, 5, 5, 1)

fit_sch <- optim(par = init_sch, fn = nll_sch, age = age, length = length, 
                 method = "L-BFGS-B", lower = lower_sch, upper = upper_sch)

cat("NLL of Schnute-Richards = ", fit_sch$value, 
    "\nL∞ = ", fit_sch$par[1], "\na = ", fit_sch$par[2],  
    "\nk = ", fit_sch$par[3], "\nc = ", fit_sch$par[4],
    "\nb = ", fit_sch$par[5], "\nσ ^ 2 = ", fit_sch$par[6] ^ 2)
```

### Conduct a comparison of the 95% confidence intervals for L among the four growth models

#### VBGF
```{r}
Lmax_vbgf_range <- seq(from = fit_vbgf$par[1] - 50,
                  to   = fit_vbgf$par[1] + 100, #set up the testing range
                  length.out = 100) #generate 100 numbers between the range

nll_vbgf_min <- fit_vbgf$value #save the most fitted NLL as the minimum NLL

Lmax_vbgf_profile_nll <- numeric(length(Lmax_vbgf_range)) #generate a empty numeric to save the results

for(i in seq_along(Lmax_vbgf_range)) {
  Lmax_vbgf_fixed <- Lmax_vbgf_range[i] #for each value in the testing range
  
  nll_vbgf_partial <- function(par){ #calculate the NLL
    K <- par[1]; t0 <- par[2]; sigma <- par[3]
    L_hat <- Lmax_vbgf_fixed * (1 - exp(-K * (age - t0)))
    L_hat[L_hat <= 0] <- 1e-6
    n <- length(length)
    n * log(sigma) + sum((log(length) - log(L_hat))^2) / (2 * sigma^2)
  }
  
  fit_vbgf_partial <- optim(par = fit_vbgf$par[2:4],
                       fn = nll_vbgf_partial,
                       method = "L-BFGS-B",
                       lower = lower_vbgf[2:4],
                       upper = upper_vbgf[2:4])
  
  Lmax_vbgf_profile_nll[i] <- fit_vbgf_partial$value
}

delta_vbgf_nll <- Lmax_vbgf_profile_nll - nll_vbgf_min #calculate the difference between each testing value and minimum

ci_vbgf_idx <- which(delta_vbgf_nll <= 1.92) #find out the value which are <= 1.92

ci_vbgf <- range(Lmax_vbgf_range[ci_vbgf_idx]) #find out the range

cat("VBGF L∞ = ", fit_vbgf$par[1], "\n95% CI: ", ci_vbgf[1], "to", ci_vbgf[2], "\n")
```

#### Gompertz
```{r}
#much the same as above
Lmax_gom_range <- seq(from = fit_gom$par[1] - 50,
                      to   = fit_gom$par[1] + 100,
                      length.out = 100)

nll_gom_min <- fit_gom$value
Lmax_gom_profile_nll <- numeric(length(Lmax_gom_range))

for(i in seq_along(Lmax_gom_range)) {
  Lmax_gom_fixed <- Lmax_gom_range[i]
  
  nll_gom_partial <- function(par){
    k2 <- par[1]; t0 <- par[2]; sigma <- par[3]
    L_hat <- Lmax_gom_fixed * exp(-exp(-k2 * (age - t0)))
    L_hat[L_hat <= 0] <- 1e-6
    n <- length(length)
    n * log(sigma) + sum((log(length) - log(L_hat))^2) / (2 * sigma^2)
  }
  
  fit_gom_partial <- optim(par = fit_gom$par[2:4],
                           fn = nll_gom_partial,
                           method = "L-BFGS-B",
                           lower = lower_gom[2:4],
                           upper = upper_gom[2:4])
  
  Lmax_gom_profile_nll[i] <- fit_gom_partial$value
}

delta_gom_nll <- Lmax_gom_profile_nll - nll_gom_min
ci_gom_idx <- which(delta_gom_nll <= 1.92)
ci_gom <- range(Lmax_gom_range[ci_gom_idx])

cat("Gompertz L∞ = ", fit_gom$par[1], "\n95% CI: ", ci_gom[1], "to", ci_gom[2], "\n")
```

#### Logistic Growth Function
```{r}
Lmax_log_range <- seq(from = fit_log$par[1] - 50,
                      to   = fit_log$par[1] + 100,
                      length.out = 100)

nll_log_min <- fit_log$value
Lmax_log_profile_nll <- numeric(length(Lmax_log_range))

for(i in seq_along(Lmax_log_range)) {
  Lmax_log_fixed <- Lmax_log_range[i]
  
  nll_log_partial <- function(par){
    k3 <- par[1]; t3 <- par[2]; sigma <- par[3]
    L_hat <- Lmax_log_fixed / (1 + exp(-k3 * (age - t3)))
    L_hat[L_hat <= 0] <- 1e-6
    n <- length(length)
    n * log(sigma) + sum((log(length) - log(L_hat))^2) / (2 * sigma^2)
  }
  
  fit_log_partial <- optim(par = fit_log$par[2:4],
                           fn = nll_log_partial,
                           method = "L-BFGS-B",
                           lower = lower_log[2:4],
                           upper = upper_log[2:4])
  
  Lmax_log_profile_nll[i] <- fit_log_partial$value
}

delta_log_nll <- Lmax_log_profile_nll - nll_log_min
ci_log_idx <- which(delta_log_nll <= 1.92)
ci_log <- range(Lmax_log_range[ci_log_idx])

cat("Logistic L∞ = ", fit_log$par[1], "\n95% CI: ", ci_log[1], "to", ci_log[2], "\n")
```

#### Schnute-Richards
```{r}
Lmax_sch_range <- seq(from = fit_sch$par[1] - 50,
                      to   = fit_sch$par[1] + 100,
                      length.out = 100)

nll_sch_min <- fit_sch$value
Lmax_sch_profile_nll <- numeric(length(Lmax_sch_range))

for(i in seq_along(Lmax_sch_range)) {
  Lmax_sch_fixed <- Lmax_sch_range[i]
  
  nll_sch_partial <- function(par){
    a <- par[1]; k <- par[2]; c <- par[3]; b <- par[4]; sigma <- par[5]
    L_hat <- Lmax_sch_fixed * (1 + a * exp(-k * age^c))^(1 / b)
    L_hat[L_hat <= 0] <- 1e-6
    n <- length(length)
    n * log(sigma) + sum((log(length) - log(L_hat))^2) / (2 * sigma^2)
  }
  
  fit_sch_partial <- optim(par = fit_sch$par[2:6],
                           fn = nll_sch_partial,
                           method = "L-BFGS-B",
                           lower = lower_sch[2:6],
                           upper = upper_sch[2:6])
  
  Lmax_sch_profile_nll[i] <- fit_sch_partial$value
}

delta_sch_nll <- Lmax_sch_profile_nll - nll_sch_min
ci_sch_idx <- which(delta_sch_nll <= 1.92)
ci_sch <- range(Lmax_sch_range[ci_sch_idx])

cat("Schnute-Richards L∞ = ", fit_sch$par[1], "\n95% CI: ", ci_sch[1], "to", ci_sch[2], "\n")
```

```{r}
#comparing the results
results_Lmax_CI <- data.frame(
  Model = c("VBGF", "Gompertz", "Logistic", "Schnute-Richards"),
  L_max = c(fit_vbgf$par[1],
            fit_gom$par[1],
            fit_log$par[1],
            fit_sch$par[1]),
  CI_lower = c(ci_vbgf[1],
               ci_gom[1],
               ci_log[1],
               ci_sch[1]),
  CI_upper = c(ci_vbgf[2],
               ci_gom[2],
               ci_log[2],
               ci_sch[2])
)

print(results_Lmax_CI)
```

### Conduct the model selection based on the AIC and quantify the plausibility of each model by using “Akaike weight (wi)
```{r}
n <- length(length)
#number of each model's parameters
p_vbgf <- 4
p_gom <- 4
p_log <- 4
p_sch <- 6
```

#### AIC
```{r}
aic_vbgf <- 2 * p_vbgf + 2 * fit_vbgf$value
aic_gom  <- 2 * p_gom  + 2 * fit_gom$value
aic_log  <- 2 * p_log  + 2 * fit_log$value
aic_sch  <- 2 * p_sch  + 2 * fit_sch$value
```

#### AICc
```{r}
aicc_vbgf <- aic_vbgf + (2 * p_vbgf * (p_vbgf + 1)) / (n - p_vbgf - 1)
aicc_gom  <- aic_gom  + (2 * p_gom  * (p_gom  + 1)) / (n - p_gom  - 1)
aicc_log  <- aic_log  + (2 * p_log  * (p_log  + 1)) / (n - p_log  - 1)
aicc_sch  <- aic_sch  + (2 * p_sch  * (p_sch  + 1)) / (n - p_sch  - 1)
```

```{r}
aicc_results <- data.frame(
  Model = c("VBGF", "Gompertz", "Logistic", "Schnute-Richards"),
  NLL = c(fit_vbgf$value, fit_gom$value, fit_log$value, fit_sch$value),
  params = c(p_vbgf, p_gom, p_log, p_sch),
  AICc = c(aicc_vbgf, aicc_gom, aicc_log, aicc_sch)
)

min_aicc <- min(aicc_results$AICc) #find out the minimum AICc value
aicc_results$delta_AICc <- aicc_results$AICc - min_aicc #calculate the delta AICc
aicc_results$Weight <- round(exp(-0.5 * aicc_results$delta_AICc) /  #calculate the weight of each model
                               sum(exp(-0.5 * aicc_results$delta_AICc)), 4)

rownames(aicc_results) <- NULL

print(aicc_results)
```

### Estimate the average model based on wi and plot all the growth curves together
```{r}
age_seq <- seq(0, max(age), by = 0.5) #set up a sequence to calculate the points of each models
```

#### VBGF
```{r}
predict_vbgf <- function(age) {
  Lmax <- fit_vbgf$par[1]
  K <- fit_vbgf$par[2]
  t0 <- fit_vbgf$par[3]
  Lmax * (1 - exp(-K * (age - t0)))
}
```

#### Gompertz
```{r}
predict_gom <- function(age) {
  Lmax <- fit_gom$par[1]
  k2 <- fit_gom$par[2]
  t0 <- fit_gom$par[3]
  Lmax * exp(-exp(-k2 * (age - t0)))
}
```

#### Logistic Growth Function
```{r}
predict_log <- function(age) {
  Lmax <- fit_log$par[1]
  k3 <- fit_log$par[2]
  t3 <- fit_log$par[3]
  Lmax / (1 + exp(-k3 * (age - t3)))
}
```

#### Schnute-Richards
```{r}
predict_sch <- function(age) {
  Lmax <- fit_sch$par[1]
  a <- fit_sch$par[2]
  k <- fit_sch$par[3]
  c <- fit_sch$par[4]
  b <- fit_sch$par[5]
  Lmax * (1 + a * exp(-k * age ^ c)) ^ (1 / b)
}
```

#### Weights
```{r}
weights <- aicc_results$Weight
names(weights) <- aicc_results$Model #name the weights results by different models' names
```

```{r}
#find out the predicted points 
L_vbgf <- predict_vbgf(age_seq)
L_gom  <- predict_gom(age_seq)
L_log  <- predict_log(age_seq)
L_sch  <- predict_sch(age_seq)
L_avg <- weights["VBGF"] * L_vbgf + weights["Gompertz"] * L_gom +
  weights["Logistic"] * L_log + weights["Schnute-Richards"] * L_sch

#open an empty plot with only the axis and title
plot(NA, NA,
     xlim = range(age_seq),
     ylim = c(min(length), max(c(length, L_vbgf, L_gom, L_log, L_sch), na.rm = TRUE)),
     xlab = "Age", ylab = "Length (mm)",
     main = "Predicted Lengths by Models")

points(age_seq, L_vbgf, col = "blue", pch = 17)
points(age_seq, L_gom,  col = "green", pch = 18)
points(age_seq, L_log,  col = "orange", pch = 15)
points(age_seq, L_sch,  col = "purple", pch = 1)
points(age, length, pch = 16, col = "grey60") #also point out the observed data

#draw the lines and the legends
lines(age_seq, L_vbgf, col = "blue")
lines(age_seq, L_gom,  col = "green")
lines(age_seq, L_log,  col = "orange")
lines(age_seq, L_sch,  col = "purple")
lines(age_seq, L_avg,  col = "red")

legend("bottomright",
       legend = c("Observrd", "VBGF", "Gompertz", "Logistic", "Schnute-Richards", "Model average"),
       col = c("grey60", "blue", "green", "orange", "purple", "red"),
       pch = c(16, 17, 18, 15, 1, -1),
       lty = c(-1, 1, 1, 1, 1, 1),
       merge = T)
```
