---
title: "homework week 11"
author: "Daniel"
date: "2025-05-03"
output: html_document
---

Dominant species
```{r}
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data/mock_dataTS")
cop_compo <- read.table("Copepod_composition.txt", header = T)
cop_den <- read.table("Cop_density.txt")
cop_sp <- read.table("copepodSPlist.txt", fill = T, sep = "\n")

sp_den <- matrix(0, nrow = length(cop_sp$V1), ncol = length(cop_den$V1))
for(i in 1:length(cop_sp$V1)){
  for(j in 1:length(cop_den$V1)){
    sp_den[i, j] <- cop_compo[i, j]/100 * cop_den[j,]
  }
}
sp_den.df <- as.data.frame(sp_den)
colnames(sp_den.df) <- colnames(cop_compo)
rownames(sp_den.df) <- cop_sp$V1
head(sp_den.df)

dominant_sp.df <- data.frame(ifelse(cop_compo >= 5, 1, 0))
dominant_species <- rownames(dominant_sp.df)[rowSums(dominant_sp.df) >= 1]

dom_sp_compo <- t(cop_compo[dominant_species, ])
colnames(dom_sp_compo) <- cop_sp[dominant_species, ]
rownames(dom_sp_compo) <- colnames(cop_compo)
dom_sp_compo.df <- as.data.frame(dom_sp_compo)
rownames(dom_sp_compo.df)[1] <- "p1"
```

Environmental data
```{r}
library(readxl)
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data/mock_dataTS")
envi_data <- read_excel("enviANDdensity.xls")
envi.df <- as.data.frame(envi_data)
rownames(envi.df) <- envi_data$station #change the row names into stations' names
envi.df$station <- NULL
rownames(envi.df)[11:34] <- c("s18", "s19", "s20", "s22", "s23", "s25", "s27", "s29", "sA",  "sB", "sC", "sD", "sE", "sF", "sG", "w22", "w23", "w25", "w27", "w29", "wA", "wB", "wC", "wD") #make sure the names are same as dominant species data
```

Apply constrained ordination
```{r}
library(vegan)
dca_y <- decorana(dom_sp_compo.df) #DCA to determine the method
summary(dca_y) #find out the axis lengths of DCA1

cop_hel <- decostand(dom_sp_compo.df, method = "hellinger") #for RDA, use hellinger method to convert original data if too many zero
rda_y <- rda(cop_hel ~ Depth + Temperature + Salinity + Fluorescence + DissolvedOxygen, data = envi.df)
cca_y <- cca(dom_sp_compo.df ~ Depth + Temperature + Salinity + Fluorescence + DissolvedOxygen, data = envi.df)

summary(rda_y) #compare both constrained inertia
summary(cca_y)

vif.cca(rda_y) #test whether those variables have multicollinearity or not
vif.cca(cca_y)

anova(rda_y, by = "term", permutations = 999) #test whether each variable is significant or not
anova(cca_y, by = "term", permutations = 999)

#because VIF result showed that temperature and dissolved  oxygen are having high multicollinearity, so test temperature only
rda_noDO <- rda(cop_hel ~ Depth + Temperature + Salinity + Fluorescence
                , data = envi.df)
cca_noDO <- cca(dom_sp_compo.df ~ Depth + Temperature + Salinity + Fluorescence
                , data = envi.df)

anova(rda_noDO, by = "term", permutations = 999)
anova(cca_noDO, by = "term", permutations = 999)

plot(rda_noDO, scaling = 1, main = "RDA (scaling 1)")
plot(rda_noDO, scaling = 2, main = "RDA (scaling 2)")
plot(cca_noDO, scaling = 1, main = "CCA (scaling 1)")
plot(cca_noDO, scaling = 2, main = "CCA (scaling 2)")
```

1.	The DCA results indicated that the length of the first axis was approximately 3.9, suggesting that both linear (RDA) and unimodal (CCA) methods should be compared using ordination plots and additional criteria.
2.	When comparing the constrained inertia between RDA and CCA, the RDA model exhibited a higher proportion of explained variation.
3.	However, the ordination diagrams showed that CCA provided better separation of the data, suggesting a clearer ecological gradient and potentially more meaningful interpretation.
4.	The VIF analysis indicated high multicollinearity between temperature and dissolved oxygen; therefore, dissolved oxygen was removed from the model to reduce redundancy.