---
title: "homework week 11"
author: "Daniel"
date: "2025-05-03"
output: html_document
---

Dominant species
```{r}
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data/mock_dataTS")
cop_compo <- read.table("Copepod_composition.txt", header = T)
cop_den <- read.table("Cop_density.txt")
cop_sp <- read.table("copepodSPlist.txt", fill = T, sep = "\n")

sp_den <- matrix(0, nrow = length(cop_sp$V1), ncol = length(cop_den$V1))
for(i in 1:length(cop_sp$V1)){
  for(j in 1:length(cop_den$V1)){
    sp_den[i, j] <- cop_compo[i, j]/100 * cop_den[j,]
  }
}
sp_den.df <- as.data.frame(sp_den)
colnames(sp_den.df) <- colnames(cop_compo)
rownames(sp_den.df) <- cop_sp$V1
head(sp_den.df)

dominant_sp.df <- data.frame(ifelse(cop_compo >= 5, 1, 0))
dominant_species <- rownames(dominant_sp.df)[rowSums(dominant_sp.df) >= 1]

dom_sp_compo <- t(cop_compo[dominant_species, ])
colnames(dom_sp_compo) <- cop_sp[dominant_species, ]
rownames(dom_sp_compo) <- colnames(cop_compo)
dom_sp_compo.df <- as.data.frame(dom_sp_compo)
rownames(dom_sp_compo.df)[1] <- "p1"
```

Environmental data
```{r}
library(readxl)
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data/mock_dataTS")
envi_data <- read_excel("enviANDdensity.xls")
envi.df <- as.data.frame(envi_data)
rownames(envi.df) <- envi_data$station #change the row names into stations' names
envi.df$station <- NULL
rownames(envi.df)[11:34] <- c("s18", "s19", "s20", "s22", "s23", "s25", "s27", "s29", "sA",  "sB", "sC", "sD", "sE", "sF", "sG", "w22", "w23", "w25", "w27", "w29", "wA", "wB", "wC", "wD") #make sure the names are same as dominant species data
```

Apply constrained ordination
```{r}
library(vegan)
dca_y <- decorana(dom_sp_compo.df) #DCA to determine the method
summary(dca_y) #find out the axis lengths of DCA1

rda_y <- rda(dom_sp_compo.df ~ Depth + Temperature + Salinity + Fluorescence +
               DissolvedOxygen, data = envi.df)
cca_y <- cca(dom_sp_compo.df ~ Depth + Temperature + Salinity + Fluorescence +
               DissolvedOxygen, data = envi.df)

summary(rda_y) #compare both constrained inertia
summary(cca_y)

plot(rda_y, scaling = 1, main = "RDA (scaling 1)")
plot(rda_y, scaling = 2, main = "RDA (scaling 2)")
plot(cca_y, scaling = 1, main = "CCA (scaling 1)")
plot(cca_y, scaling = 2, main = "CCA (scaling 2)")
```

1. The DCA results indicated that the length of the first axis was approximately 3.9, suggesting that both linear (RDA) and unimodal (CCA) methods should be compared using ordination plots and additional criteria.
2. When comparing the constrained inertia between RDA and CCA, the RDA model exhibited a higher proportion of explained variation.
3. However, the ordination diagrams showed that CCA provided better separation of the data, suggesting a clearer ecological gradient and potentially more meaningful interpretation.
4. In the CCA scaling 1 plot, the distances among sites are well-resolved, allowing comparison of community composition differences and their alignment with environmental gradients. However, species labels are highly overlapped and difficult to interpret.
5. In contrast, the scaling 2 plot presents clearer associations between species and environmental variables—e.g., *Oithona* *similis* aligns with the direction of increasing dissolved oxygen—yet site distribution is more cluttered, limiting its usefulness for site-based analysis.