---
title: "homework week 03"
author: "Daniel"
date: "2025-03-10"
output: html_document
---
```{r}
#setup
setwd("C:/Users/dan91/Rstudio/statistic-in-ecology/stat_data")
library(readxl)
f_c_den <- read_excel("mock_dataTS/enviANDdensity.xls")
```
# Compute the mean and SE(mean) for the fish and copepod densities
# Plot the histogram of bootstrapped means with bootstrap 1000 times.

## Normal theory
### mean

```{r}
Normal_f_mean <- mean(f_c_den$`FishDensity (ind./1000m3)`)
Normal_c_mean <- mean(f_c_den$`CopepodDensity ind./m3)`)
print(Normal_f_mean)
print(Normal_c_mean)
```

### SE

```{r}
Normal_f_SE <- sd(f_c_den$`FishDensity (ind./1000m3)`) / sqrt(length(f_c_den$`FishDensity (ind./1000m3)`))
Normal_c_SE <- sd(f_c_den$`CopepodDensity ind./m3)`) / sqrt(length(f_c_den$`CopepodDensity ind./m3)`))
print(Normal_f_SE)
print(Normal_c_SE)
```

## Bootstrap
### mean

```{r}
B <- 999 #bootstrap 1000 times
n <- length(f_c_den$`FishDensity (ind./1000m3)`) #number of the data
bootstrap_fish_mean <- numeric(B) #generate a empty numeric with 1000 space to store the results
bootstrap_copepod_mean <- c(Normal_c_mean, numeric(B))

set.seed(123)
for(i in 1:B){
  r <- ceiling(runif(n, 1, n)) #generate n number of random integer to represent the bootstrap sample
  resample_f_mean <- f_c_den$`FishDensity (ind./1000m3)`[r] #select the samples into resample
  resample_c_mean <- f_c_den$`CopepodDensity ind./m3)`[r]
  
  bootstrap_fish_mean[i + 1] <- mean(resample_f_mean) #calculate mean of every resample
  bootstrap_copepod_mean[i + 1] <- mean(resample_c_mean)
}
par(mfrow = c(1, 2))
hist(bootstrap_fish_mean, breaks = 50)
hist(bootstrap_copepod_mean, breaks = 50)
```

### SE(mean)

```{r}
bootstrap_fish_SE <- sd(bootstrap_fish_mean)
print(bootstrap_fish_SE)

bootstrap_copepod_SE <- sd(bootstrap_copepod_mean)
print(bootstrap_copepod_SE)
```

# Compute the median and bootstrapped SE(median) for the fish and copepod densities. 
# Plot the histogram of bootstrapped medians with bootstrap 1000 times.

## Normal theory
### fish median

```{r}
f_median <- median(f_c_den$`FishDensity (ind./1000m3)`)
```

### copepod median

```{r}
c_median <- median(f_c_den$`CopepodDensity ind./m3)`)
```

## Bootstrap
### fish SE(median)

```{r}
bootstrap_fish_median <- numeric(B)
bootstrap_copepod_median <- numeric(B)

set.seed(123)
for(i in 1:B){
  r <- ceiling(runif(n, 1, n))
  resample_f_median <- f_c_den$`FishDensity (ind./1000m3)`[r]
  resample_c_median <- f_c_den$`CopepodDensity ind./m3)`[r]

  bootstrap_fish_median[i + 1] <- median(resample_f_median)
  bootstrap_copepod_median[i + 1] <- median(resample_c_median)
}
```

### SE(median)

```{r}
par(mfrow = c(1, 2))
Bootstrap_f_SE_median <- sd(bootstrap_fish_median)
print(Bootstrap_f_SE_median)
hist(bootstrap_fish_median, breaks = 50)

Bootstrap_c_SE_median <- sd(bootstrap_copepod_median)
print(Bootstrap_c_SE_median)
hist(bootstrap_copepod_median, breaks = 50)
```


# Plot fish (dependent) v.s copepod (independent) and the regression line. 
# Compute the regression coefficients for fish = b0+b1*copepod and bootstrapped SE(b0) and SE(b1). 
# Plot the histogram of bootstrapped b0 and b1 with bootstrap 1000 times.

```{r}
x <- cbind(1, f_c_den$`CopepodDensity ind./m3)`) #combine 1 into the matrix to calculate the intercept, which is b0
y <- f_c_den$`FishDensity (ind./1000m3)`
b <- solve(t(x) %*% x) %*% t(x) %*% y #use the equation: b = (x'x)^-1x'y, to calculate b0 and b1
b0 <- b[1] #since add 1 into the matrix, the first element is b0
b1 <- b[2] #the second one is b1

y_hat <- x %*% b  #prediction
residuals <- y - y_hat  #residual
sigma_hat <- sqrt(sum(residuals^2) / (n - 2))  # standard deviation of residual

Normal_SE_b1 <- sigma_hat / sqrt(sum((f_c_den$`CopepodDensity ind./m3)` - mean(f_c_den$`CopepodDensity ind./m3)`))^2))
Normal_SE_b0 <- sigma_hat * sqrt(1/n + (mean(f_c_den$`CopepodDensity ind./m3)`)^2) / sum((f_c_den$`CopepodDensity ind./m3)` - mean(f_c_den$`CopepodDensity ind./m3)`))^2))

print(Normal_SE_b0)
print(Normal_SE_b1)

par(mfrow = c(1, 1))
plot(f_c_den$`CopepodDensity ind./m3)`, y, main = "Fish v.s. Copepod", xlab = "copepod density", ylab = "fish density")
abline(b0, b1, col = "lightblue", lwd = 1.5) 
```

### Bootstrap b0 and b1

```{r}
bootstrap_b0 <- numeric(B)
bootstrap_b1 <- numeric(B)

set.seed(123)
for(i in 1:B){
  r <- ceiling(runif(n, 1, n))
  
  resample_x <- x[r,] #resample x to do bootstrap
  resample_y <- y[r] #also resample y
  
  bootstrap_b <- solve(t(resample_x) %*% resample_x) %*% t(resample_x) %*% resample_y #solve the equation again to get b
  
  bootstrap_b0[i + 1] <- bootstrap_b[1]
  bootstrap_b1[i + 1] <- bootstrap_b[2]
}

#calculate the SE of b0 and b1
bootstrap_SE_b0 <- sd(bootstrap_b0)
bootstrap_SE_b1 <- sd(bootstrap_b1)

#calculate the mean of b0 and b1
Bootstrap_b0_mean <- mean(bootstrap_b0)
Bootstrap_b1_mean <- mean(bootstrap_b1)

#plot the histogram of both
par(mfrow = c(1, 2))
hist(bootstrap_b0, breaks = 50)
hist(bootstrap_b1, breaks = 50)
```

# Compute the mean and stand error of the mean for the fish and copepod density (all data points) respectively using Jackknife. 
# Plot the histogram of Jackknife means.

```{r}
jackknife_f_mean <- numeric(n)
jackknife_c_mean <- numeric(n)

for(i in 1:n){ #for each sample of jackknife, minus one data in each sample and then calculate the mean
  jackknife_f_mean[i] <- mean(f_c_den$`FishDensity (ind./1000m3)`[-i])
  jackknife_c_mean[i] <- mean(f_c_den$`CopepodDensity ind./m3)`[-i])
}
Jackknife_f_mean <- mean(jackknife_f_mean)
Jackknife_c_mean <- mean(jackknife_c_mean)

#calculate SE of jackknife by the equation: 
jackknife_f_SE_mean <- sqrt((n - 1) / n * sum((jackknife_f_mean - mean(jackknife_f_mean)) ^ 2))
jackknife_c_SE_mean <- sqrt((n - 1) / n * sum((jackknife_c_mean - mean(jackknife_c_mean)) ^ 2))

#plot the histogram
par(mfrow = c(1, 2))
hist(jackknife_f_mean, breaks = 50)
hist(jackknife_c_mean, breaks = 50)
```

# Compute the regression coefficients for fish = b0+b1*copepod and Jackknife SE of b0 and b1.
# Plot the histogram of Jackknife b0 and b1.

```{r}
jackknife_b0 <- numeric(n)
jackknife_b1 <- numeric(n)

for(i in 1:n){
  jackknife_x <- cbind(1, f_c_den$`CopepodDensity ind./m3)`[-i]) #same as above to add 1 with the matrix to calculate b0
  jackknife_y <- f_c_den$`FishDensity (ind./1000m3)`[-i]
  
  jackknife_b <- solve(t(jackknife_x) %*% jackknife_x) %*% t(jackknife_x) %*% jackknife_y
  
  jackknife_b0[i] <- jackknife_b[1]
  jackknife_b1[i] <- jackknife_b[2]
}

#calculate the SE of b0 and b1
jackknife_SE_b0 <- sqrt((n - 1) / n * sum((jackknife_b0 - mean(jackknife_b0)) ^ 2))
jackknife_SE_b1 <- sqrt((n - 1) / n * sum((jackknife_b1 - mean(jackknife_b1)) ^ 2))

Jackknife_b0_mean <- mean(jackknife_b0)
Jackknife_b1_mean <- mean(jackknife_b1)

#plot the histogram
par(mfrow = c(1, 2))
hist(jackknife_b0, breaks = 50)
hist(jackknife_b1, breaks = 50)
```

# Compare the estimates for Q1 and Q2 obtained from normal theory, bootstrap, and jackknife.

```{r}
mean <- c(Normal_f_mean, Normal_c_mean)
median <- c(f_median, c_median)
f_c_compare <- data.frame(mean, median, row.names = c("fish density", "copepod density"))
f_c_compare

b0_b1 <- data.frame(b0, b1)
b0_b1

fish_SE_mean <- c(Normal_f_SE, bootstrap_fish_SE, jackknife_f_SE_mean)
copepod_SE_mean <- c(Normal_c_SE, bootstrap_copepod_SE, jackknife_c_SE_mean)
SE_b0 <- c(Normal_SE_b0, bootstrap_SE_b0, jackknife_SE_b0)
SE_b1 <- c(Normal_SE_b1, bootstrap_SE_b1, jackknife_SE_b1)
method_compare <- data.frame(fish_SE_mean, copepod_SE_mean, SE_b0, SE_b1, row.names = c("Normal", "Bootstrap", "Jackknife"))
method_compare

par(mfrow = c(1, 1))
plot(f_c_den$`CopepodDensity ind./m3)`, y, main = "Fish v.s. Copepod", xlab = "copepod density", ylab = "fish density")
abline(b0, b1, col = "lightblue", lwd = 2)
abline(Bootstrap_b0_mean, Bootstrap_b1_mean, col = "aquamarine", lwd = 2)
abline(Jackknife_b0_mean, Jackknife_b1_mean, col = "brown2", lwd = 2, lty = 2)
legend("bottomright", legend = c("Normal", "Bootstrap", "Jackknife"), 
       col = c("lightblue", "aquamarine", "brown2"), lty = c(1, 1, 2))
```